Adaptive Safe Strong Screening for Efficient Lasso-Type Problems Optimization over the Solution Path

### 2021-01-16 main.tex
Edits to 3.4

One promising technique to address this challenge is feature screening. The lasso model solution is sparse in the sense that most coefficients will be exactly zero; for these ``inactive'' coefficients we do not need their associated features.  In theory, if we knew which coefficients were nonzero at each value of $\lambda$, we could read into memory only those features and leave the remaining features on disk. In reality, we cannot know this prior to fitting the model -- however, through the clever use of feature screening rules, we can greatly reduce the number of times inactive features are read into memory, thereby minimizing the computational burden. There are many screening techniques where the solution obtained after screening will still be the global optimum given it exists, and in this paper we will only focus on this type of screening. In what follows, we refer to features left on disk as ``discarded'' features, although it is worth clarifying that this is not a permanent decision.  In particular, a feature can be discarded at one value of $\lambda$ but included in the list of potentially active features at other values of $\lambda$.

@@

Typical practice when fitting lasso models is to solve for $\hat{\beta}(\lambda)$ along a decreasing sequence of $K+1$ values of $\lambda$: $\lambda_0 > \lambda_1 > \cdots > \lambda_K$.  As we will discuss shortly, we set $\lambda_0=\max_j|x_j^Ty/n|$ throughout; this ensures that $\hat{\beta}(\lambda_0)=0$, making it a natural point from which to start.

@@

Compared to strong rules, safe rules are never violated. Features can be discarded safely, with no need for post-covergence checking, as the coefficients are guaranteed to be zero by the rule. We will mainly focus on enhanced dual polytope projections (EDPP) rules\citep{wang2013lasso} because they outperform other known safe rules and are easy to incorporate into our method.

@@

In this paper, we will only focus on screening methods that never falsely discard an active feature. There are also almost sure screening methods where probability of false discards is positive but tends to 0 under certain conditions, such as Sure Independence Screening \citep{Fan2008}.

@@

The SEDPP rule \eqref{eq:sedpp} from Theorem 2.2 allows us to screen features at $\lambda_{k+1}$ using the solution at $\lambda_k$ as a reference. However, since the grid of $\lam$ values is arbitrary, the rule with the same reference can be just as easily applied to screen solutions at $\lam_{k+2}, \lam_{k+3}, \ldots$, until some $\lam_{k+B}$, where we update the reference to the current solution at $\lam_{k+B}$ and screen the following $\lam$ values.  The difference is that in typical SEDPP usage, we would update the reference solution at every step of path; here, we are proposing to re-use solutions for several values of $\lam$, only updating the reference when it is computationally beneficial to do so.  We refer to this modification as \emph{adaptive EDPP}.  Adaptive EDPP is still a safe rule, but it is neither sequential nor one-step.  A similar idea would be to use the reference solution in batches, each one being used for a fixed number of $\lam$ steps.  However, since an appropriate batch size is difficult to determine ahead of time, and also because the ideal size changes along the path, we find that an adaptive approach is better. \pb{Do we have evidence for this?}

@@

\begin{algorithm}[H]
    \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Pre}{Pre-compute}\SetKwInOut{Initialize}{Initialize}
    \SetAlgoLined
    \Input{$\{x_j\}_{j=1}^p, \{x_j^Ty\}_{j=1}^p,  \lambda_{0},...,\lambda_{K}$}
    \BlankLine
    \Initialize{reference = $\lambda_0$}

@@

In this section we will fill in line 11 in Algorithm 1. We will develop an algorithm to adaptively determining when to update the reference for screening, by evaluating screening power at previous $\lambda$-values since the last update. Our goal is to minimize the computation cost, so we will first analyse computation complexity for each part in Algorithm 1. Then we will propose an adaptive algorithm that can minimize the entire computation cost.

@@

This complexity analysis will only consider calculations with $O(np)$ complexity. The rest calculations, those with $O(p)$ complexity for example, are not only faster in order, but also consume less memory and thus usually do no need out-of-core computing. Calculation of $\{x_j^Ty\}_{j=1}^p$ is needed only once for the whole solution path, so it would be neglected too. We will assume that strong rule screening is always powerful in the sense that $|\mathcal{R}_b|=o(p)\approx0$, but safe rule screening may not. We also assume that post-convergence check will mostly pass in the first iteration and thus is only needed approximately once per $\lambda$-value.

Under these assumptions, there are two dominating parts in the algorithm within each update. First, the computation of $\{x_j^Tr(\lambda_k)\}_{j=1}^p$ at the time of updating the reference, which requires $np$ multiplications. Then, for each $\lambda$-value $\lambda_{k+b}$, the post-convergence check requires approximately $|\mathcal{S}_{k+b}\setminus\mathcal{R}_{k+b}|\approx|\mathcal{S}_{k+b}|$ KKT condition checks and each KKT condition check requires $n$ multiplications. These sum up to $n\sum_{i=k}^{k+B}|\mathcal{S}_k|$ multiplications, where $k+B$ is the start of the next update. Combining these two parts, we will have the average cost per $\lambda$-value in the update $\lambda_{k+1},...,\lambda_{k+B}$ to be

\begin{equation}
    \label{eq:cost}
    \bar{T}_k(B) \approx n\frac{p}{B}+n\frac{\sum_{b=1}^B|\mathcal{S}_{k+b}|}{B}.
\end{equation}

The first term is the average cost for safe screening. Delaying next update (increasing $B$) means reusing calculation at the previous update more times for the safe screening and thus reducing the average cost. The second term is the average cost for KKT condition checking. If the old reference is reused too many times, safe rule screening becomes less powerful and the number of KKT condition checks required ($|\mathcal{S}_{k+b}|$) increases.

To better illustrate these two effects, we considered the most naive and simple updating scheme: update the reference after every batch of $B$ $\lambda$ values. The following plot illustrates the effect of update delay $B$ on $|\mathcal{S}_{k+b}|$ in a real data set solved over a path of $K=100$ $\lambda$-values. The y axis is the proportion of discarded features, calculated as $1-|\mathcal{S}_{k+b}|/p$.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.6]{plots/batchsizes.pdf}    \caption{Proportion of features rejected by screening rules vs index of $\lambda$-values for different update delay.}
    \label{fig:3.4.1}
\end{figure}

BEDPP rule is the special case of adaptive EDPP that never updates ($B=K$). When the reference does not get updated for a long time, the safe rule screening will eventually fail to discard any features and make the KKT condition checking cost increase to its maximum $np$. SEDPP rule is the special case that updates at every $\lambda$ value ($B=1$). It remains very powerful along the path, making the KKT condition checking cost small, but the safe screening itself is costly and takes $np$ time per $\lambda$-value. Adaptive EDPP can strike a balance between BEDPP and SEDPP. In this example, simply using fixed update delay $B=10$ can reduce the safe screening cost by 90\% compared to SEDPP and reduce the KKT checking cost by 50\% to more than 90\% compared BEDPP.

This example also shows the timing of updates should be decided carefully and adaptively. For the first 10 $\lambda$-values, the safe rule stays powerful, which means the old reference is still exploitable. We may want delay the update to reduce average safe screening cost with little harm to the KKT condition checking cost. For the last 20 $\lambda$-values however, the screening power drops to 0 quickly, which means the reference becomes useless quickly. Updating more frequently can greatly reduce the KKT condition checking cost with a relatively small additional safe screening cost. Moreover, the performance of adaptive EDPP rule also depends on many factors such as the data, the range of $\lambda$-values and spacing between $\lambda$-values. These suggest that simple updating rules like fixed batch updating cannot not be consistently efficient and we want a method that can automatically determine whether to update given current performance.

@@

From previous example, we assume the average cost is a convex function of update delay $B$, which decreases first and then after some optimal value increases again. If we want to minimize the average cost per $\lambda$-value, we need to update when using the old reference for one more $\lambda$-value will increase the average cost. That is, we should update the reference to $\lambda_{k+B}$ when $\bar{T}_k(B+1)>\bar{T}_k(B)$. For practice, we will update the reference to $\lambda_{k+B}$ when $\bar{T}_k(B)>\bar{T}_k(B-1)$ instead, because $\bar{T}_k(B+1)$ cannot be calculated without doing screening at $\lambda_{k+B+1}$. These two conditions should be close and thus this substitution should also be close to optimal. Combined with \eqref{eq:cost}, the latter condition will approximately be:

@@

This updating condition is a greedy condition that only considers the cost up to the current $\lambda$ value. In practice, we cannot know about future $\lambda$ values so this is the most reasonable condition. In fact, it provides a robust strategy towards three most common scenarios. First, if the screening power stays high, meaning $\mathcal{S}_{k+b}$ is small for all $b$, then the left-hand-side of \eqref{eq:crit} will be small and thus at this point we will not update to further exploit the old reference. Second, if the screening power has been high for a while but dropped suddenly, meaning $\mathcal{S}_{k+B}$ is large but $\mathcal{S}_{k+b}$ are mostly small for $b=1,2,...,B-1$, then the left-hand-side of \eqref{eq:crit} will be large and we will update the useless old reference to a helpful new one. Last, if the screening power drops dramatically right after the last update, meaning $\mathcal{S}_{k+b}$ are close to $p$ for $b=2,3,...,B$, then the left-hand-side of \eqref{eq:crit} will be approximately $p-|\mathcal{S}_1|<p$. We will not update although old reference is useless, because if we did, the new reference would become useless quickly as well.

This updating condition fills in the condition on line 11 in Algorithm 1 and forms a complete pathwise lasso algorithm with adaptive hybrid screening.

>>>>

One promising technique to address this challenge is feature screening. The lasso model solution is sparse in the sense that most coefficients will be exactly zero; for these ``inactive'' coefficients we do not need their associated features.  In theory, if we knew which coefficients were nonzero at each value of $\lambda$, we could read into memory only those features and leave the remaining features on disk. In reality, we cannot know this prior to fitting the model -- however, through the clever use of feature screening rules, we can greatly reduce the number of times inactive features are read into memory, thereby minimizing the computational burden. There are many screening techniques where the solution obtained after screening is guaranteed to be the global optimum if one exists, and in this paper we focus only on this type of screening. In what follows, we refer to features left on disk as ``discarded'' features, although it is worth clarifying that this is not a permanent decision.  In particular, a feature can be discarded at one value of $\lambda$ but included in the list of potentially active features at other values of $\lambda$.

@@

Typical practice when fitting lasso models is to solve for $\hat{\beta}(\lambda)$ along a decreasing sequence of $K+1$ values of $\lambda$: $\lambda_0 > \lambda_1 > \cdots > \lambda_K$.  As we discuss shortly, we set $\lambda_0=\max_j|x_j^Ty/n|$ throughout; this ensures that $\hat{\beta}(\lambda_0)=0$, making it a natural point from which to start.

@@

Compared to strong rules, safe rules are never violated. Features can be discarded safely, with no need for post-covergence checking, as the coefficients are guaranteed to be zero by the rule. We mainly focus on enhanced dual polytope projections (EDPP) rules\citep{wang2013lasso} because they outperform other known safe rules and are easy to incorporate into our method.

@@

In addition to the deterministic screening methods we have described, there are also probabilistic screening rules such as Sure Independence Screening \citep{Fan2008}. In this paper, we consider only screening methods that never falsely discard an active feature and that are thus guaranteed to converge to the global solution. This is not the case for probabilistic screening approaches, which construct screening rules such that the probability of incorrectly discarding active features tends to 0 asymptotically under certain conditions, but for any finite sample size, have a positive probability of converging to an inferior solution.

@@

The SEDPP rule \eqref{eq:sedpp} from Theorem 2.2 allows us to screen features at $\lambda_{k+1}$ using the solution at $\lambda_k$ as a reference. However, since the grid of $\lam$ values is arbitrary, the rule with the same reference can be just as easily applied to screen solutions at $\lam_{k+2}, \lam_{k+3}, \ldots$, until some $\lam_{k+B}$, where we update the reference to the current solution at $\lam_{k+B}$ and screen the following $\lam$ values.  The difference is that in typical SEDPP usage, we would update the reference solution at every step of path; here, we are proposing to re-use solutions for several values of $\lam$, only updating the reference when it is computationally beneficial to do so.  We refer to this modification as \emph{adaptive EDPP}.  Adaptive EDPP is still a safe rule, but it is neither sequential nor one-step.  A similar idea would be to use the reference solution in batches, each one being used for a fixed number of $\lam$ steps.  However, since an appropriate batch size is difficult to determine ahead of time, and also because the ideal size changes along the path, we find that an adaptive approach is better (see Section~\ref{sec:adaptive}).

@@

\begin{algorithm}[H]
  \label{alg:path}
    \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Pre}{Pre-compute}\SetKwInOut{Initialize}{Initialize}
    \SetAlgoLined
    \Input{$\{x_j\}_{j=1}^p, \{x_j^Ty\}_{j=1}^p,  \lambda_{0},...,\lambda_{K}$}
    \BlankLine
    \Initialize{reference = $\lambda_0$}

@@

In this section we will fill in line 11 in Algorithm~\ref{alg:path}. We develop an algorithm to determine when to update the reference for screening. This decision is guided by evaluating screening power at previous $\lambda$ values, and is therefore adaptive, updating the reference more or less frequently as we continue along the solution path. Our goal is to minimize the computation cost, so we will first analyse computation complexity for each part in Algorithm 1. Then we will propose an adaptive algorithm that can minimize the entire computation cost.

@@

This complexity analysis will only consider calculations with $O(np)$ complexity. The remaining calculations, those with $O(p)$ complexity for example, are not only of lower order, but also consume less memory and thus usually do no require out-of-core computing. Calculation of $\{x_j^Ty\}_{j=1}^p$ is needed only once for the whole solution path; we ignore this as well as its computational cost is the same across all approaches. We assume that strong rule screening is always powerful in the sense that its set of non-discarded features, $\mathcal{R}_{k+b}$, satisfies $|\mathcal{R}_{k+b}|=o(p) \approx 0$, but that safe rule screening, with corresponding set $\mathcal{S}_{k+b}$, may not be. We also assume that post-convergence check will usually pass in the first iteration and thus is only needed approximately once per $\lambda$.

Under these assumptions, there are two dominating parts in the algorithm within each update. First, the computation of $\{x_j^Tr(\lambda_k)\}_{j=1}^p$ at the time of updating the reference, which requires $np$ multiplications. Then, for each $\lambda_{k+b}$, the post-convergence check requires approximately $|\mathcal{S}_{k+b} \setminus \mathcal{R}_{k+b}| \approx |\mathcal{S}_{k+b}|$ KKT condition checks, each of which requires $n$ multiplications. These sum to $n\sum_{i=k}^{k+B}|\mathcal{S}_k|$ multiplications, where $k+B$ is the start of the next update. Combining these two parts, the average cost per $\lambda$ in the block $\lambda_{k+1},...,\lambda_{k+B}$ is
\begin{equation}
    \label{eq:cost}
    \bar{T}_k(B) \approx n\frac{p}{B}+n\frac{\sum_{b=1}^B|\mathcal{S}_{k+b}|}{B},
\end{equation}
where the first term represents the cost of updating the safe screening rule and the second term the cost of KKT checking. Delaying the next update (increasing $B$) means reusing calculations additional times, thus reducing the cost of safe screening (first term). However, reusing an old reference reduces the power of the safe rule, thereby increasing the cost of KKT checking (second term).

To better illustrate these two effects, we considered a simple updating scheme: update the reference after every batch of $B$ $\lambda$ values. Figure~\ref{fig:batchsizes} illustrates the effect of the batch size $B$ on $|\mathcal{S}_{k+b}|$ in a real data set solved over a path of $K=100$ $\lambda$-values. The vertical axis is the proportion of discarded features, $1-|\mathcal{S}_{k+b}|/p$.

\begin{figure}[ht]
    \centering
    \includegraphics[scale = 0.6]{plots/batchsizes.pdf}    \caption{Proportion of features rejected by screening rules vs index of $\lambda$-values for different update delay.}
    \label{fig:batchsizes}
\end{figure}

BEDPP is the special case of adaptive EDPP that never updates ($B=K$). When the reference is not updated, safe rule screening will eventually fail to discard any features and the KKT condition checking cost will increase to its maximum, $np$. SEDPP is the special case that updates at every $\lambda$ value ($B=1$). It remains very powerful along the path, making the KKT condition checking cost small, but safe screening updates themselves are costly and require $np$ operations at every $\lambda$-value. Adaptive EDPP can strike a balance between BEDPP and SEDPP. In this example, simply using a fixed batch size $B=10$ can reduce the safe screening cost by 90\% compared to SEDPP and reduce the KKT checking cost by 50\% to more than 90\% compared BEDPP \pb{I find this confusing \ldots by 50\% to more than 90\%?  What does that mean?}.

This example also illustrates the appeal of choosing updates in an adaptive manner. For the first 10 $\lambda$ values, the safe rule stays powerful, which means the old reference is still useful. In this portion of the solution path, we want to delay the update in order to reduce the cost of safe screening, with little harm to the cost of KKT condition checking. For the last 20 $\lambda$ values however, the reference quickly becomes useless. In this portion of the solution path, we want to update more frequently in order to reduce the KKT condition checking cost. Moreover, the performance of adaptive EDPP depends on many factors such as the data, the range of $\lambda$-values and spacing between $\lambda$-values. Thus, simply updating the EDPP reference after a fixed batch size will not be as efficient as a method that can adaptively determine whether or not to update.

@@

As explained in the previous section, it is reasonable to assume that the average cost, as a function of the update delay $B$, decreases at first, then reaches an optimal value before it begins increasing again. A reasonable approach to minimizing the average cost, therefore, is to update the reference at the point where reusing it one additional time would increase the average cost. That is, we should update the reference at $\lambda_{k+B}$ if $\bar{T}_k(B+1)>\bar{T}_k(B)$. In practice, we will instead update the reference to $\lambda_{k+B}$ when $\bar{T}_k(B)>\bar{T}_k(B-1)$, because $\bar{T}_k(B+1)$ cannot be calculated without having already carried out screening at $\lambda_{k+B+1}$. The two conditions should be close and thus this substitution should be near optimal. Combined with \eqref{eq:cost}, the latter condition is approximately

@@

This updating condition is a greedy condition that only considers the cost up to the current $\lambda$ value. This is unavoidable in practice, as we cannot know about future $\lambda$ values. The resulting adaptive rule, \eqref{eq:crit}, provides a robust strategy that addresses the three most common scenarios. First, if the screening power stays high, meaning $\mathcal{S}_{k+b}$ is small for all $b$, then the left-hand-side of \eqref{eq:crit} will be small and thus at this point we will not update to further exploit the old reference. Second, if the screening power has been high for a while but dropped suddenly, meaning $\mathcal{S}_{k+B}$ is large but $\mathcal{S}_{k+b}$ are mostly small for $b=1,2,...,B-1$, then the left-hand-side of \eqref{eq:crit} will be large and we will update the useless old reference to a helpful new one. Last, if the screening power drops dramatically immediately after the last update, meaning $\mathcal{S}_{k+b}$ are close to $p$ for $b=2,3,...,B$, then the left-hand-side of \eqref{eq:crit} will be approximately $p-|\mathcal{S}_1|$. This is less than $p$, so we will again not update.  Although the old reference is useless, there is little point in calculating a new reference because if we did, it would quickly become useless as well.

We use this updating condition at line 11 in Algorithm~\ref{alg:path} to form a complete pathwise lasso algorithm with adaptive hybrid screening.
